---
output:
  pdf_document: default
  html_document: default
  word_document: default
---
STAA552 Handout: Valve Replacement
========================================================
#### Jay Breidt, Department of Statistics, Colorado State University

**Agresti Table 4.5.**
In this study, 109 heart valve replacement patients were followed until death, dropout, or the end of the study. Age group (`Age`) and valve type (`ValveType`) were recorded along with number of deaths (`Deaths`) and time at risk (`TimeAtRisk`) in months for each cell.
```{r}
Deaths <- c(4, 1, 7, 9)
TimeAtRisk <- c(1259, 2082, 1417, 1647)
ValveType <- c("aortic", "mitral", "aortic", "mitral")
Age <- c("<55", "<55", "55+", "55+")
d <- data.frame(ValveType, Age, Deaths, TimeAtRisk)
d
```
Let $A_i = 1$ if `Age = 55+` and 0 otherwise, and $M_i = 1$ if `ValveType =
mitral` and 0 otherwise. 

For cells $i = 1,2,3,4$, model the data as
$$
Y_i\sim\mbox{independent Poisson}(t_i\mu(x_i))
$$
where $t_i$ is the known `TimeAtRisk` (in months) and
$$
\log(\mbox{E}[Y_i])=\log t_i+\sum_{j=0}^p\beta_jx_{ij}.
$$
First fit the  the saturated model, with one parameter for each of the four
cells, just out of curiosity:
```{r}
Saturated <- glm(Deaths ~ Age * ValveType, offset = log(TimeAtRisk), 
                 family = poisson(link = log))
# notice that Age + ValveType + Age * ValveType is the saturated model.
summary(Saturated)
```
Now look at the additive GLM
$$
\log(\mbox{E}[Y_i ]) = \log t_i +\beta_0 + \beta_1A_i + \beta_2 M_i.
$$
```{r}
Full <- glm(Deaths ~ Age + ValveType, offset = log(TimeAtRisk), 
            family = poisson(link = log))
summary(Full)
1 - pchisq(Full$deviance, df = 1)
```
The comparison of the additive GLM to the Saturated model shows no strong
evidence of lack-of-fit at level 0.05. Estimated numbers of deaths and rates
per month are then
```{r}
d <- data.frame(d, Expected_Deaths = exp(predict(Full)), 
                Rates_per_Month = exp(predict(Full)) / TimeAtRisk)
d
```
These are the values under "Log Link"" in Table 4.6 of Agresti.

What we really want to know is whether the type of valve makes a difference.  Accordingly, we try to simplify the model further:
```{r}
Reduced <- glm(Deaths ~ Age, offset = log(TimeAtRisk), family = poisson(link = log))
summary(Reduced)
anova(Reduced, Full)
```
What should we conclude?

