---
output:
  pdf_document: default
  html_document: default
  word_document: default
---
STAA552 Handout:  Alligator Food Preference 
========================================================
#### Jay Breidt, Department of Statistics, Colorado State University



**From Agresti 8.1.2.** $N = 219$ alligators were trapped from four Florida lakes (Hancock, Oklawaha, Trafford, George) and their stomach contents were examined. Primary food source for the $i$th alligator was classified as $Y_i\in\{\mbox{fish,bird,invertebrate,reptile,other}\}$.
In addition, $M_i = 1$ for male, 0 for female and $S_i = 1$ for small ($\le 2.3$m) and
0 for large were recorded.
```{r, warning = FALSE}
library(nnet)    # has the function multinom
library(vcdExtra)# contains the alligator data set
data(Alligator)  # get the alligator data from the vcdExtra package
names(Alligator)
head(Alligator)
dim(Alligator)
```
Notice that instead of one record per alligator, the data are grouped, so that there are 7 small, male, Hancock alligators that prefer fish. We will account for this grouping in the modeling by using a `weights` statement.

Model the data as 
$$
Y_i\sim\mbox{independent Multinomial}\left(1;\left[\pi_F(x),\pi_B(x),\pi_I(x),\pi_R(x),\pi_O(x)\right]\right)
$$
for $i=1,2,\ldots,219$. Using the baseline-category approach, choose `bird` as baseline and fit several models.
```{r}
# Saturated model.
fitS <- multinom(food ~ lake * size * sex, data = Alligator, weights = count)	
# Null model.
fit0 <- multinom(food ~ 1, data = Alligator, weights = count)  			
fit1 <- multinom(food ~ sex, data = Alligator, weights = count) 			
fit2 <- multinom(food ~ size, data = Alligator, weights = count)  			
fit3 <- multinom(food ~ lake, data = Alligator, weights = count)  		
fit4 <- multinom(food ~ size + lake, data = Alligator, weights = count) 		
fit5 <- multinom(food ~ size + lake + sex, data = Alligator, weights = count) 	
```
Look at a particular model fit:
```{r}
summary(fit5)
```
Reproduce Table 8.2, column 2 of Agresti.
```{r}
fit0$deviance - fitS$deviance # null deviance
fit1$deviance - fitS$deviance # sex only
fit2$deviance - fitS$deviance # size only
fit3$deviance - fitS$deviance # lake only
fit4$deviance - fitS$deviance # lake + size
fit5$deviance - fitS$deviance # sex + lake + size
```
The small difference between model 4 (dropping `sex`) and model 5 suggests
that the model can be simplified. The drop-in-deviance test is
```{r}
fit4$deviance-fit5$deviance
```
to be compare to $\chi^2_4$; dropping `sex` saves one parameter in each of the $J-1$ non-baseline categories, hence 4 degrees of freedom.  Yes, we can simplify. 

Now collapse over `sex` to reproduce Table 8.2, column 6 of Agresti:
```{r}
# Saturated model:
fitS <- multinom(food ~ lake * size, data = Alligator, weights = count)	
# Null model:
fit0 <- multinom(food ~ 1, data = Alligator, weights = count)  			
fit1 <- multinom(food ~ size, data = Alligator, weights = count)
fit2 <- multinom(food ~ lake, data = Alligator, weights = count)
fit3 <- multinom(food ~ size + lake, data = Alligator, weights = count)
fit0$deviance - fitS$deviance # null
fit1$deviance - fitS$deviance # size only
fit2$deviance - fitS$deviance # lake only
fit3$deviance - fitS$deviance # size + lake
```
Think about degrees of freedom for these models.  The null model says there are no differences among lakes or sizes, so only one probability distribution over the five classes (`fish`, `invert`, `other`, `reptile` and `bird`) is necessary.  This requires four parameters, since the probabilities sum to one, and `bird` as baseline is the category without a separate parameter:
```{r}
coef(fit0)
```
The saturated model has 32 parameters: 4 each for 4 lakes times 2 size classes.  Hence the null model has $32 - 4 = 28$ residual degrees of freedom.  

The `size` only model has two distributions, one for small alligators and one for large alligators (default), hence 8 parameters and 24 residual df:
```{r}
coef(fit1)
```
Similiarly, the `lake` only model has four distributions, one for each lake (with Lake George as default), hence uses 16 df and has 16 residual df: 
```{r}
coef(fit2)
```

Finally, the `size + lake` model is additive: it uses 20 parameters (either 16 lake parameters to get 4 lake distributions, which are then shifted by 4 size parameters;  or 8 size parameters to get 2 size distributions, which are then shifted by 12 lake parameters) and hence has 12 residual df: 
```{r}
coef(fit3)
```
Among these non-saturated models, `fit3` is the largest model and has the smallest residual deviance (as it must).  Is it better than the next best model, `fit2`? Or could we simplify the model? Drop-in-deviance is 
```{r}
fit2$deviance - fit3$deviance
```
on $16 - 12 = 4$ df, with p-value 
```{r}
1 - pchisq(fit2$deviance - fit3$deviance, df = 4)
```
so reject the proposed simplification.

If you are a small alligator, what is your food preference relative to a large alligator?

Finally, do some probability calculations, both with "predict" and via direct computations.  Recall that `bird` is the baseline and look at the default probabilities, which would be for a large alligator in Lake George: 
```{r}
tmp <- c(0, coef(fit3)[, 1]) # 0 for bird and Lake George coefficients.
exp(tmp) / sum(exp(tmp))
predict(fit3,
        type = "probs",
        newdata = data.frame(size = "large", lake = "George"))
```

But we want Lake Hancock, so we shift the Lake George coefficients.  We want the default, large alligator, so we don't use the `small` coefficients in column 2:  

```{r}
tmp <- c(0, coef(fit3)[, 1] + coef(fit3)[, 3])
exp(tmp) / sum(exp(tmp))
predict(fit3,
        type = "probs",
        newdata = data.frame(size = "large", lake = "Hancock"))
```
The above computations reproduce the results on Agresti, page 297.


What if we wanted a small alligator in Lake George?  

What if we wanted a small alligator in Lake Trafford?
